Running an open experiment: transparency and reproducibility in soil and ecosystem

### 2016-04-18 manuscript_openexperiment/Open_experiment.Rmd
Incorporate VLB feedback on open experiment draft

Researchers in soil and ecosystem science, and almost every other field, as increasingly being pushed--by funders, journals, governments, and themselves--to increase transparency and reproducibility of their work. A key part of this effort is a move towards *open data* as a way to fight post-publication data loss, improve data and code quality, enable powerful meta- and cross-disciplinary analyses, and increase trust in, and the efficiency of, publicly-funded research. Here we describe a recent "open experiment", in which we documented every aspect of a soil incubation experiment online, making all raw data, scripts, diagnostics, final analyses, and manuscripts publicly available in real time. We found that using tools such as version control, issue tracking, and open-source statistical software improved data integrity, accelerated our team's communication and productivity, and ensured transparency. There are many avenues to improve scientific reproducibility and data availability, of which is this only one example, and it is not an approach suited for every experiment or situation. Nonetheless, we encourage the communities in our respective fields to consider its advantages, and to lead rather than follow with respect to scientific reproducibility, transparency, and data availability.

@@

As data and code have become increasingly intertwined, the availability of the latter has become a fundamental problem as well. Specialized modeling groups have worked to improve reproducibility and archiving practices {Thornton, 2005 #2310}. More generally, scientists in all fields are increasingly building and using software in their work, though often without strong training in this area {Wilson, 2014 #4337}. In addition to open-source data analysis languages such as R {R Development Core Team, 2016 #4314}, scientific workflow systems such as Kepler (https://kepler-project.org) or Taverna (http://www.taverna.org.uk) record information about the data processing, analytical process and decisions, and statistical analysis. Providing open code does not magically produce bug-free code, mistake-free analyses, or instantly better science {Easterbrook, 2014 #4137}. But it does encourage authors to invest the time upfront to clean up their code, data, and documentation when a paper is written, rather than deferring this task--often until key details have been forgotten, if not forever.

@@

Data loss hits ecosystem, soil, and global change ecology particularly hard, as climate changes make ecological data effectively irreproducible {Wolkovich, 2012 #3303}: we can never remeasure exactly the same system state. This results in a critical role for syntheses and meta-analyses {Ellison, 2010 #4363}{Cassey, 2006 #4349}, which depend on the existence of documented protocols and (ideally raw) data. A subtler data loss issue is the "file drawer problem" {Rosenthal, 1979 #2291}, where unpublished but potentially valuable data are lost forever. Scientists' use of strong and consistent data curation practices {Hart, 2015 #4292}{Rüegg, 2014 #3886} can mitigate the problem, but all our anecdotal experience and quantitative studies {Vines, 2014 #4130} suggest that in the long term, data cannot be reliably preserved by individual researchers.

This in turns means that the role of established, structured data and code repositories is critical. These provide a much-needed improvement on "Supplementary Information" sections accompanying journal articles, which have become *de facto* repositories of data, but often have inconvenient formats (e.g. PDF), poorly searchability, use restrictions, and uncertain long-term availability {Molloy, 2011 #4366;Neylon, 2012 #4367}. Most journals do not have expertise in data storage and management {Cassey, 2006 #4349}. In contrast, the best repositories provide easy data uploading, immediate assignment of data digital object identifiers (DOIs), and long-term stability and availability. They do not, however, eliminate all risks: for example, the genomics sciences have a long history of requiring sequences to be deposited, but there are multiple repositories of varying curatorial levels, data quality, and formats, causing longer-term problems {Lagesen, 2010 #4369}.

@@

Transparency is no magic panacea ensuring repeatability and reproducibility {Easterbrook, 2014 #4137}, and may raise genuine concerns about, for example, protecting scientists from harassment {Lewandowsky, 2016 #4342}. Nor will open science by itself fundamentally enhance public trust in science, as there are still many social and political challenges to overcome {Grand, 2012 #4370}.

>>>>

Researchers in soil and ecosystem science, and almost every other field, are being pushed--by funders, journals, governments, and themselves--to increase transparency and reproducibility of their work. A key part of this effort is a move towards *open data* as a way to fight post-publication data loss, improve data and code quality, enable powerful meta- and cross-disciplinary analyses, and increase trust in, and the efficiency of, publicly-funded research. Here we describe a recent "open experiment", in which we documented every aspect of a soil incubation experiment online, making all raw data, scripts, diagnostics, final analyses, and manuscripts publicly available in real time. We found that using tools such as version control, issue tracking, and open-source statistical software improved data integrity, accelerated our team's communication and productivity, and ensured transparency. There are many avenues to improve scientific reproducibility and data availability, of which is this only one example, and it is not an approach suited for every experiment or situation. Nonetheless, we encourage the communities in our respective fields to consider its advantages, and to lead rather than follow with respect to scientific reproducibility, transparency, and data availability.

@@

As data and code have become increasingly intertwined, the availability of the latter has become a fundamental problem as well. Specialized modeling groups have worked to improve reproducibility and archiving practices {Thornton, 2005 #2310}. More generally, scientists in all fields are increasingly building and using software in their work, though often without strong training in this area {Wilson, 2014 #4337}. In addition to open-source data analysis languages such as R {R Development Core Team, 2016 #4314}, scientific workflow systems such as Kepler (https://kepler-project.org) or Taverna (http://www.taverna.org.uk) record information about the data processing, analytical process and decisions, and statistical analysis. Providing open code does not magically produce bug-free code, mistake-free analyses, or instantly better science {Easterbrook, 2014 #4137}. But it does encourage authors to invest the time upfront to clean up their code, data, and documentation when a paper is written, rather than deferring this task--often until key details have been forgotten, if not forever. This then presents opportunities for real-time peer review of both code and data.

@@

Data loss hits ecosystem, soil, and global change ecology particularly hard, as climate changes make ecological data effectively irreproducible {Wolkovich, 2012 #3303}: we can never remeasure exactly the same system state. This results in a critical need for syntheses and meta-analyses {Ellison, 2010 #4363}{Cassey, 2006 #4349}, which depend on the existence of documented protocols and (ideally raw) data. A subtler data loss issue is the "file drawer problem" {Rosenthal, 1979 #2291}, where unpublished but potentially valuable data are lost forever. Scientists' use of strong and consistent data curation practices {Hart, 2015 #4292}{Rüegg, 2014 #3886} can mitigate the problem, but all our anecdotal experience and quantitative studies {Vines, 2014 #4130} suggest that in the long term, data cannot be reliably preserved by individual researchers.

This in turns means that the role of established, structured data and code repositories is critical. These provide a much-needed improvement on "Supplementary Information" sections accompanying journal articles, which have become *de facto* repositories of data, but often have inconvenient formats (e.g. PDF), use restrictions, and uncertain long-term availability {Molloy, 2011 #4366;Neylon, 2012 #4367}. Most journals do not have the necessary and forward-thinking expertise in data storage and management {Cassey, 2006 #4349}. In contrast, the best repositories provide easy data uploading, immediate assignment of data digital object identifiers (DOIs), and long-term stability and availability. They do not, however, eliminate all risks: for example, the genomics sciences have a long history of requiring sequences to be deposited, but there are multiple repositories of varying curatorial levels, data quality, and formats, causing longer-term problems {Lagesen, 2010 #4369}.

@@

By itself transparency does not guarantee repeatability and reproducibility {Easterbrook, 2014 #4137}, and it may raise genuine concerns about, for example, protecting scientists from harassment {Lewandowsky, 2016 #4342}. Nor will open science by itself fundamentally enhance public trust in science, as there are still many social and political challenges to overcome {Grand, 2012 #4370}.
